# Methods {#methods}

```{r ch2-setup, include = F}

library(plyr)
library(dplyr)
library(ggplot2)
library(tmap)
library(ggmap)

```
## Chapter Overview

This chapter gives an overview of the clinical study which underpins this thesis, and the laboratory and computational procedures used in analysis. Further details are given in the individual chapters, where necessary.

## Study site

### Malawi

Malawi is a country of 17.5 million people in South-Eastern Africa (2018 Census, Malawi national statistical office). It is one of the poorest countries in the world: it is a low income country under the World Bank classifcation, with a 2017 Gross National Income (GNI) per capita of \$320 in US dollars, and was ranked 171st of 189 countries in 2017 by the human development index (HDI), a composite statistic of life expectancy, education and income per capita indicators (UNDP human development reports). In 2010, 71% of the population was estimated to survive on less than \$1.90 per day (World Bank). Life expectancy at birth in 2017 was 63 years, and though significant progress is being made, neonatal and under-5 mortality remains high at 23 and 55 per 1000 live births, respectively (world bank). The population is largely rural (83% in 2017, World Bank), with a young population (44% under the age of 15, 2017, World Bank) and high fertility rate (4.6, 2016 data, World Bank). Malaria is endemic, and there is an ongoing generalised HIV epidemic: adult HIV prevalance (age 15-49) was estimated to be 9.6% in 2017 (UNAIDS). It is classed by the WHO as a high-TB/high-HIV burdern country, with an estimated TB incidence rate of 131 (95% CI 70-210) cases per 100000 population per year. It has a subtropcial climate, with three main seasons: a warm wet season from November to April, a cooler dry winter period from May to August and a hot dry period from September to October. Blantyre city, the location of the study in this thesis, is the second city of Malawi with a population of 995000 in 2018 (Malawi ONS). It is located in Blantyre district in the Shire highlands at an altitude of 1000m (Figure \@ref(tab:plot-malawi-map)).

```{r plot-malawi-map, warning = F,  echo = F, message = F, fig.cap = "Malawi, showing administrative boundaries (North, Central, and South regions), Lilongwe, the capital city and Blantyre, the study location. Source: openstreetmap.org, used under Creative Commons Attribution ShareAlike 2.0 licence CC-BY-SA"}


afbox <- bbox <- c(left = -21.22, right = 55.02, top= 17.17, bottom = -37.75)

bbox <- c(left = 25.840, right = 42.737, top= -8.277, bottom = -18.167)
bbox2 <- c(left = 25.955, right = 38.717, top= -8.277, bottom = -18.167)
#get_stamenmap(bbox2, zoom = 8) -> m
get_stamenmap(bbox2, zoom = 7, maptype = "terrain-background") -> m2

get_stamenmap(afbox, zoom = 4, maptype= "terrain-background") -> a

#ggmap(m2) 

cities <- data.frame(city = c("Blantyre", "Lilongwe"), lat = c(-15.7667,-13.9626), lon = c(35.0168, 33.7741))


ggmap(a) + geom_polygon(data = mal2, aes(x= long, y = lat, group = group), fill = "black") +
  theme_inset() +
  geom_rect(xmin = bbox2["left"], xmax = bbox2["right"], ymin = bbox2["bottom"], ymax = bbox2["top"], fill = NA, colour = "black" )-> p1

ggmap(m2) +
  geom_path(data = mal2, aes(x= long, y = lat, group = group), alpha = 0.8) + 
  geom_point(data = cities, aes(x = lon, y = lat)) + 
  geom_text(data = cities, aes(x = lon, y = lat,label = city,  vjust = -0.5, hjust = 0.5, size = 7, fontface = "bold")) +
  theme_void() + theme(legend.position = "none") +  inset(grob = ggplotGrob(p1), xmin = 26.5, xmax = 31.5, ymin = -13, ymax = -9)


```

### Queen Elizabeth Central Hospital

### Participating Laboratories

## Clinical Study

### Summary

### Entry Criteria

### Study Team

### Study Visits and Patient Sampling

## Diagnostic Laboratory Procedures

## Quality Assurance

## Data Storage

## Molecular Methods

## Bioinformatics

## Ethical Approval and Consent
