# Sepsis in Blantyre, Malawi
 
```{r ch4-setup, include = F}

library(plyr)
library(dplyr)
library(ggplot2)
library(tmap)
library(ggmap)

wd <- "chapter_4/"
output = "latex"

# generate figures
source("chapter_4/generate_consort_diagram.R")
source("final_cleaning_scripts/load_and_clean_followup_and_enroll.R")
source("other_scripts/summary_strings.R")

```

## Chapter overview

## Methods

blah blah



## Results

### Study population

Figure \@ref(fig:consort-diag) shows flow through the study. 225 patients were recruited in under 20 month, between 19th February 2017 and 2nd October 2018. In total, 4 patients (2%) were lost to follow up over the 180-day study period; 5 patients (2%) withdrew; and 7 patients (3%) transferred out of the study area before 180 days. Four of the five patients who withdrew gave a reason for their wish to withdraw, all that they no longer wished the inconvenience of bing involved in the study. 15/225 (7%) patients had their final study visit before 180 days, and so were not included in the 180-day outcome analysis.

```{r consort-diag, echo = F, fig.cap = "Study recruitment and follow up."}

knitr::include_graphics(paste0(wd, "/figures/consort_diagram.pdf"))

```

```{r sep-t1-demog, echo = F, fig.cap = "Study recruitment and follow up."}

e1 <- subset(enroll, arm == 1)



```


Table - demographics

Table - presentation

Table - health seeking behaviour

### Aetiology

Table

Figure to show crossover

### Treatment

Table:
Time to antimicrobials
Time to fluid
Amount of fluid

### Outcome

Table - 28 and 90 day mortality

Figure - KM survival curve

Logistic regression - determinants of 28 day mortality

Morbidity - 
