# Longitudinal ESBL-E carriage in Malawian adults in health and disease

\chaptermark{ESBL-E carriage}

```{r ch4-setup, include = F}

library(plyr)
library(tidyverse)
library(knitr)
library(kableExtra)
library(grid)
library(ggpubr)
library(pheatmap)
library(RColorBrewer)
library(scales)
library(lubridate)
library(survival)
library(ggsci)
library(survminer)
library(glue)


wd <- "chapter_5/"
output = "latex"
T <- TRUE

specify_decimal <- function(x, k) trimws(format(round(x, k), nsmall=k))


# generate figures
#source("chapter_5/make_consort_diagram.R")
#source("final_cleaning_scripts/generate_visits_df.R")
source("final_cleaning_scripts/load_and_clean_followup_and_enroll_labelled.R")
source("other_scripts/summary_table_functions.R")
source("final_cleaning_scripts/make_composite_hivstatus_variable.R")
source("final_cleaning_scripts/load_and_clean_hourly.R")
#source("final_cleaning_scripts/load_and_clean_bloods.R")
#source("final_cleaning_scripts/load_and_clean_aetiol.R")
#source("final_cleaning_scripts/load_and_clean_hourly.R")
source("final_cleaning_scripts/load_and_clean_upto72.R")
source("final_cleaning_scripts/load_and_clean_post72.R")
source("final_cleaning_scripts/load_and_clean_hosp_oc.R")
#source("final_cleaning_scripts/load_and_clean_time_to_ab.R")
#source("final_cleaning_scripts/load_and_clean_fluid_hr1_to_6.R")


# lims




#panel data helper functions
source("other_scripts/panel_data_helpers/expand_covariates.R")
source("other_scripts/panel_data_helpers/sort_out_tb_rx_on_discharge.R")
source("other_scripts/panel_data_helpers/shuffle_a_in_b2.R")
source("other_scripts/panel_data_helpers/strip_post_dropout_rows.R")
source("other_scripts/panel_data_helpers/extract_covariate_exposure.R")
source("other_scripts/panel_data_helpers/collapse_covariates.R")
source("other_scripts/panel_data_helpers/ditch_everything_after_first_1.R")
source("other_scripts/panel_data_helpers/mstate_helper_functions.R")

names(enroll)[names(enroll) == "data_date"] <- "enroll_date"
followup <- merge(followup, select(enroll, pid, arm, enroll_date), all.x = T)
followup$t <- followup$data_date - followup$enroll_date

# and load lims
source("final_cleaning_scripts/load_and_clean_lims.R")

```

## Chapter Overview

## Introduction and chapter aims

## Methods

## Results

### Study population

In total, 425 participants were recuited to the  study between 19th February 2017 and 2nd October 2018; 225 participants with sepsis (arm 1), 100 inpatients without antimicrobial exposure at baseline (arm 2) and 100 community members (arm 3). Flow of participants through the study is shown in Figure \@ref(fig:ESBL-consort-diag). It was often challenging to collect stool samples from participants but 87% (1416/1631) eligible patient-visits resulted in the collection of a stool sample. Drop out from the study and failure to collect stool samples were similar in arm 1 and 2 and with no apparent systematic bias, but both drop out and missing samples were less frequent in arm 3 (Figure \@ref(fig:ESBL-prop-samples)A). There was siginifcant variation in the timing of stool sample collecion, with a distribution around the ostensible collation day (Figure \@ref(fig:ESBL-prop-samples)B).

The baseline characetristics of the enrolled participants are shown in Table \@ref(tab:dassim-2-demog). There were some important differences between the arms of the study: despite matching on age and sex, antimicrobial-unexposed participants were older. They were also less likely to be HIV-infected than participants with sepsis (13% [12/89] of those with known HIV status were HIV-infected versus 67% [143/213] with sepsis), and less likely to have been treated for TB. Sepsis participants were more likely to have recieved antimicroials or been hospitalised in the previous 4 weeks. In the community arm of the study, there were a high proportion of participants (60% [60/100]) with an unknown HIV status, and there were some differences in toilet facilities, water sources, cooking fuel and presence of animals at home across the three groups. 

```{r ESBL-prop-samples, echo = F, warning = F, message = F,fig.scap="Missing samples and variation in sample collection time", out.extra='', fig.cap = "A: Missing stool samples stratified by arm and visit. Bar height at a given visit represents the number of eligible participants, coloured by successful sample collection (blue) or failure to collect a sample (red). B: Distribution of actual day of sample collection for ostensible day 7, 28, 90 and 180 samples showing considerable variation.",  fig.height = 4, fig.width=8, fig.align = 'center'}

sampls <- read.csv("chapter_5/ESBL_carriage_recruitment.csv", stringsAsFactors = F)

sampls$a_missed <- sampls$eligible - sampls$collected

sampls %>% select(-eligible) %>% pivot_longer(-c(arm, visit)) %>% ggplot(aes(as.factor(visit), value, fill = name)) + geom_col() + facet_wrap(~arm, scales = 'free') + xlab("Visit day") + ylab("n") + theme_bw() + theme(legend.title = element_blank(), legend.position = "top") + scale_fill_manual(labels = c("No sample", "Sample collected"), values = hue_pal()(2)) -> p1


visit_names <- c(
  `1` = "Day 7",
  `2` = "Day 28",
  `3` = "Day 90",
  `4` = "Day 180"
)

followup %>% filter(arm != 4, foutcome == 1) %>% ggplot(aes(as.numeric(t), group = as.factor(d2visit))) + geom_histogram(bins = 60) + facet_wrap(~ d2visit, scale = 'free_y', ncol = 1, strip.position = "right", labeller = as_labeller(visit_names)) + xlim(c(0,300)) + theme_bw() + xlab("Time post enrollment (days)") + ylab("n") -> p2

ggarrange(ggarrange(NULL,p1,NULL, ncol = 1, nrow = 3, heights = c(0.15,1,0.15), labels = c(NA,"A",NA)),p2, ncol = 2, nrow = 1, labels = c(NA,"B"))

```


```{r ESBL-consort-diag, echo = F, fig.scap="ESBL carriage study recruitment and follow up.", out.extra='', fig.cap = "Study recruitment and follow up. At each time point \\textit{eligible participants} refers to participants who are known to be alive and have not withdrawn from the study by that time point, and \\textit{samples collected} refers to patients from whom a stool sample was sucessfully collected for that visit.",  out.height = '100%', fig.align = 'center'}

knitr::include_graphics(paste0(wd, "/figures/ESBLstudy_flow_diagram.odf"))

```


\renewcommand{\arraystretch}{0.8}

```{r dassim-2-demog, echo = F, warning=F, message = F}

enroll <- filter(enroll, arm != 4)
enroll$art.time <- as.numeric((enroll$enroll_date - enroll$hivartstart) /(365.25/12))
enroll$hivart[enroll$hivart != "5A" & !is.na(enroll$hivart)] <- "ART regimen: other"
enroll$animalskept[enroll$animalskept == "Other"] <- "Elsewhere"
enroll$keep.other[enroll$keep.cattle == "Yes" | enroll$keep.sheep == "Yes" | enroll$keep.mules == "Yes"] <- "Yes"
enroll.sum <- dplyr::select(enroll, arm,calc_age, ptsex, hivstatus,recieved_prehosp_ab,pmhxrechospital,tbstatus, tbongoing, hivonart, hivart, art.time, hivcpt, 
                  tobaccoyn, alcoholyn,  highestedu,
                 job, housholdadultsno, householdchildno, toilet , watersource, watertreated,
                 electricityyn, fuel, 
                 keepanim, keep.poultry,  keep.goats,
                 keep.dogs, keep.cattle,  keep.other)

enroll.sum$arm <- as.character(enroll.sum$arm)

enroll.sum %>% select_if(is.character) %>% 
  pivot_longer(-arm) %>% 
  group_by(name, arm, value) %>% 
  tally() %>% 
  pivot_wider(names_from = arm, values_from = n, , values_fill = list(n = 0) ) %>% filter(!is.na(value)) %>% group_by(name) %>% 
  mutate(sum1 = sum(`1`), 
         sum2 = sum(`2`), 
         sum3 = sum(`3`),
         str1 = glue('{`1`}/{sum1} ({specify_decimal(`1`*100/sum1, 0)}%)'),
         str2 = glue('{`2`}/{sum2} ({specify_decimal(`2`*100/sum2, 0)}%)'),
         str3 = glue('{`3`}/{sum3} ({specify_decimal(`3`*100/sum3, 0)}%)')
  ) -> t1

apply(t1[3:8], 1, function(x) fisher.test(matrix(x, ncol = 2))$p.value) -> t1$p
t1$tot <- t1$`1` + t1$`2` + t1$`3`
t1$n <- t1$sum1 + t1$sum2 + t1$sum3
t1$Total <- t1 %>% glue_data('{tot}/{n} ({specify_decimal(tot*100/n, 0)}%)')
t1$totalprop <- t1$tot /t1$n
t1 <- filter(t1, value != "No")

enroll.sum$arm <- as.numeric(enroll.sum$arm)

enroll.sum %>% select_if(is.numeric) %>% 
  pivot_longer(-arm) %>% 
  group_by(name) %>% mutate(p = kruskal.test(value ~ arm)$p.value,
                            median.t = median(value, na.rm = T),
                                    lqr.t = quantile(value, 0.25, na.rm = T),
                                    uqr.t = quantile(value, 0.75, na.rm = T),
                                    Total = glue('{specify_decimal(median.t,1)} ({specify_decimal(lqr.t,1)}-{specify_decimal(uqr.t,1)})')) %>%
  group_by(name, arm) %>% summarise(median = median(value, na.rm = T),
                                    lqr = quantile(value, 0.25, na.rm = T),
                                    uqr = quantile(value, 0.75, na.rm = T),
                                    strz = glue('{specify_decimal(median,1)} ({specify_decimal(lqr,1)}-{specify_decimal(uqr,1)})'),
                                    p = unique(p),
                                    Total = unique(Total)) %>% select(name, arm, strz, p, Total) -> t1.1 

dcast(t1.1, name+ p + Total ~ arm, value.var = "strz") -> t1.1
names(t1.1)[4:6] <- c("str1", "str2", "str3")
  bind_rows(dplyr::select(t1, name, value,str1, str2, str3, p, Total, totalprop), t1.1) -> t1
    rm(t1.1)  
    
    t1$value[grepl("keep.", t1$name) & t1$name != "keepanim"] <- t1$name[grepl("keep.", t1$name)& t1$name != "keepanim"]
     t1$name[grepl("keep.", t1$name)& t1$name != "keepanim"] <- "which.anim"
    ord <- c("calc_age", "ptsex", 
             "hivstatus", "tbstatus", "tbongoing",
             "hivonart", "art.time", "hivart", "hivcpt",
             "recieved_prehosp_ab", "pmhxrechospital",
             "tobaccoyn","alcoholyn", 
             "highestedu", "job", 
             "toilet",
             "watersource",
             "watertreated",
             "householdchildno",
             "housholdadultsno",
             "electricityyn",
             "fuel",
             "keepanim",
             "which.anim"
             )
        
       t1$name <- factor(t1$name, levels = ord)  
       t1[order(t1$name, -t1$totalprop),] -> t1
       
# tweaky tweaky
       
       
    t1$name <- as.character(t1$name)


t1$value[t1$name == "householdchildno"] <- "Children"
t1$value[t1$name == "housholdadultsno"] <- "Adults"
t1$name[t1$name == "householdchildno"] <- "No. household members"
t1$name[t1$name == "housholdadultsno"] <- "No. household members"
t1$value[t1$name == "calc_age"] <- "Age (yr)"
t1$value[t1$name == "art.time"] <- "Months on ART"

t1 <- subset(t1, value != "Female")
t1$value[t1$name == "ptsex"] <- "Male sex"
t1$name[t1$name %in% c("calc_age", "ptsex")] <- "Demographics"

# art variables
t1$value[t1$value %in% c("Reactive", "Non reactive", "Unknown")] <- 
  paste0("HIV ", str_to_title(t1$value[t1$value %in% c("Reactive", "Non reactive", "Unknown")]))

t1$value[t1$name == "hivonart"] <- "Current ART*"

t1$value[t1$name == "hivcpt"] <- "Current CPT\\textsuperscript{\\dag}"
t1$value[t1$value == "5A"] <- "ART regimen: EFV/3TC/TDF"
t1 <- subset(t1, value != "ART regimen: other")

#t1$levels[t1$variable == "art.time"] <- "Months on ART"
t1$name[t1$name == "hivstatus"] <- "HIV/TB status"
t1$name[grepl("art", t1$name)] <- "ART status*"

t1$name[grepl("cpt", t1$name)] <- "ART status"
# tb variables
t1$value[t1$name == "tbstatus"] <- "Ever treated for TB"
t1$value[t1$name == "tbongoing"] <- "Of those, current TB treatment"
t1$name[grepl("tb", t1$name)] <- "HIV/TB status"

t1$value[t1$name == "recieved_prehosp_ab"] <- "Antibiotics"
t1$value[t1$name == "pmhxrechospital"] <- "Hospitalised"

t1$name[t1$name == "recieved_prehosp_ab"] <- "Healthcare exposure last 4wk"
t1$name[t1$name == "pmhxrechospital"] <- "Healthcare exposure last 4wk"




# tobacco use
t1$value[grepl("tobacco", t1$name)] <- paste0(t1$value[grepl("tobacco", t1$name)], " tobacco" )
t1$name[grepl("tobacco", t1$name)] <- "Tobacco/alcohol use"

# alcohol use
t1$value[t1$name == "alcoholyn"] <- "Current alcohol"
t1$name[grepl("alcohol", t1$name)] <- "Tobacco/alcohol use"

# education
t1$name[grepl("highestedu", t1$name)] <- "Education"

# employment 
t1$name[grepl("job", t1$name)] <- "Employment"

# toilet

t1$name[grepl("toilet", t1$name)] <- "Toilet facilities"

# toilet
t1$value[t1$name == "watertreated"] <- "Treat water with chlorine"

t1$name[grepl("water", t1$name)] <- "Main water source"

# electricity
t1$value[t1$name == "electricityyn"] <- "Electricity available in house"
t1$name[grepl("electric", t1$name)] <- "Electricty"

t1$name[grepl("fuel", t1$name)] <- "Main cooking fuel"

# animals

#t1$levels[grepl("keep", t1$variable)] <- t1$variable[grepl("keep", t1$variable)]

sub("keep.", "",t1$value) -> t1$value
t1$name[t1$name == "keepanim" | t1$name == "which.anim"] <- "Animals at home?"
t1$value[t1$name == "Animals at home?"] <- str_to_title(t1$value[t1$name == "Animals at home?"])
t1$value[t1$name == "Animals at home?" & t1$value == "Yes"] <- "Any animal"

str_to_title(t1$levels[grepl("keep", t1$variable)] ) -> t1$levels[grepl("keep", t1$variable)]
#t1$levels[t1$levels == "Nim"] <- "Any animal"
names(t1)[1] <- "variable"
kstring <- make_kable_rowgroup_string(t1)


sub("%", "\\\\%", t1$str1) -> t1$str1
sub("%", "\\\\%", t1$str2) -> t1$str2
sub("%", "\\\\%", t1$str3) -> t1$str3
sub("%", "\\\\%", t1$Total) -> t1$Total
boldrows <- t1$p < 0.05

t1$p <- specify_decimal(t1$p, 3)
t1$p[t1$p == "0.000"] <- "<0.001"

t1 <- ungroup(t1)
boldrows <- which(boldrows %in% TRUE)

kable(select(t1,value, str1, str2, str3, p,Total), "latex", booktabs =T, 
      row.names = F, longtable = T, escape = F, col.names = c("Variable","Sepsis", "Inpatient", "Community", "p", "Total"), 
      caption = "Participant Characteristics" ) %>%  kable_styling(font_size = 9, full_width = F,latex_options = c("scale_down")) %>%
  pack_rows(index = kstring) %>%
  row_spec(boldrows, bold = T) %>%
  footnote(general = "ART = Antiretroviral therapy, CPT = Co-trimoxazole preventative therapy, EFV: Efavirenz, 3TC: Lamivudine, TDF: Tenofovir. Numeric values are median (IQR)) unless otherwise stated. P-values are to assess for differened across the three groups: Fisher's exact test across the groups for categorical variable, and Kruskal-Wallace test for continuous variables.", symbol = c("ART status includes HIV reactive only as denominator", "Missing CPT data for two participants. ") ,threeparttable = T, fixed_small_size = T) %>% landscape()




                                               
```

### Exposures druring the study period

Exposures to antimicrobials and hospitalisation of the cohort are shown in Figures \@ref(fig:ESBL-study-exp) and \@ref(fig:ESBL-a2-time-to-ab) and Table \@ref(tab:exposure-tab). Antimicrobial-unexposed inpatients (Arm 2 participants) had a shorter length of hospital stay than participants with sepsis (median [IQR] 2 [2-7] versus 5 [2-10] days, p = 0.002 by Kruskal-Wallace test). Five of the 100 Arm 2 participants were taking co-trimoxazole preventative therapy (CPT) at baseline, 18 received further courses of antimicrobials during the study period, and two were started on TB therapy. Some participants received combinations of these therapies, so in total 23% (23/100) Arm 2 participants received an antibacterial during the study period, most commonly within 30 days following enrollment (Figure \@ref(fig:ESBL-a2-time-to-ab)), and most commonly ceftriaxone (Table \@ref(tab:exposure-tab)).

Both antimicrobial exposure and hospitalisation were unusal in the community cohort; 7% (7/100) community (Arm 3) participants were taking CPT and one received a 5-day course of amoxicillin meaning that 8% (8/100) Arm 3 participants received an antibacterial during the study period. In addition one Arm 3 participant was hospitalised for 1 day in the study period. No Arm 3 participant received any TB therapy, and no Arm 2 or 3 participants received any antimalarial or antifungal therapy during the study period.

The most commonly receievd antibacterial by Arm 1 participants - those with sepsis - (apart from co-trimoxazole and TB therapy) was ceftrixaxone by some distance with 998 participant-days of expsoure in 189 participants during the study period, and a median 5 (IQR 3-7) day course on admission. Ciprofloxacin and amoxicillin were also commonly received, with 61 participants receiving 398 participant-days of exposure to ciprofloxacin with a median 7 (IQR 5-7) day course, and 39 participants receiving 235 participant-days of exposure to amoxicillin with a median 5 (IQR 5-7) day course.  Because of the chronic nature of the therapy, the greatest exposure (in terms of participant-days) were to co-trimoxazole and TB therapy, by an order of magnitude (Table \@ref(tab:exposure-tab)).

```{r ESBL-study-exp, echo = F, warning = F, message = F,fig.scap="Hospital and antibacterial exposure of participants", out.extra='', fig.cap = "Hospital and antibacterial exposure of participants expressed as (A) proportion of participants in the sepsis (arm 1) and inpatient antimicrobial-unexposed (arm 2) groups who are hospitalised and exposed to the most commonly received antibacterials, as a function of time and (B) participant-days of exposure to the most commonly received antibacterials (excluding TB therapy and co-trimoxazole) across the three arms of the study over the whole study period.  ",  fig.height = 8, fig.width=8, fig.align = 'center'}

read.csv("data/longit_covariate_data.csv") -> df.covs

df.covs <- merge(df.covs, select(enroll, pid, arm))


df.covs %>% group_by(pid) %>% do(uncompress_covariates(.)) -> outlong
df.covs.cpt <- merge(df.covs, select(enroll, pid, hivcpt), all.x = T)

df.covs.cpt$hivcpt <- as.numeric(df.covs.cpt$hivcpt == "Yes")
df.covs.cpt$cotri[df.covs.cpt$hivcpt == 1] <- 1
df.covs.cpt <- select(df.covs.cpt, -hivcpt)

df.covs.cpt %>% group_by(pid) %>% do(uncompress_covariates(.)) -> outlong.cpt

abs.lookup <- c(
  "Amoxicillin",
  "Gentamicin",
  "Azithromycin",
  "TB therapy",
  "Penicillin",
  "Fluconazole",
  "Ceftriaxone",
  "Amphotericin",
  "Quinine",
  "Chloramphenicol",
  "LA",
  "Ciprofloxacin",
  "Co-amoxiclav",
  "Artesunate",
  "Cotrimoxazole",
  "Clindamycin",
  "Doxycycline",
  "Erythromycin",
  "Aciclovir",
  "Flucloxacillin",
  "Metronidazole",
  "Streptomycin"
)

names(abs.lookup) <- outlong %>% ungroup() %>% select(-c(pid, assess_type, died, hosp, arm)) %>% names 

abs.lookup <- c(abs.lookup, "Hospitalised", "Other AB")
names(abs.lookup) <- c(names(abs.lookup)[1:(length(abs.lookup) - 2)], "hosp", "other")

#other.antibac <- c("genta", "azithro","benzy", "coamo","clinda", "doxy","erythro", "fluclox", "metro", "strepto")

#apply(outlong.cpt[other.antibac],1,sum) -> outlong.cpt$other


arm_names <- c(
  `1` = "Arm 1: Sepsis",
  `2` = "Arm 2: Inpatient",
  `3` = "Arm 3: Community"
)


outlong.cpt %>%
  ungroup() %>% 
  select(-c(pid, died)) %>% filter(arm != 3) %>%
  pivot_longer(-c(assess_type, arm)) %>%
  group_by(arm, assess_type, name)%>%
  summarise(n = n(), x = sum(value == 1), prop = x/n)  %>% 
  filter(name %in% c("cefo","amoxy","hosp", "cipro", "cotri", "tb")) %>%
  mutate(name = recode(name, !!!abs.lookup)) %>%
  ggplot(aes(assess_type,prop, group = name, color = name, linetype = name)) +
  geom_line(alpha = 0.8, size = 0.75) + 
  facet_wrap(~ arm, ncol = 1, , labeller = as_labeller(arm_names)) + 
  theme_bw() + 
  scale_x_continuous(breaks = c(0,5,10,15,20,25,30), limits = c(0,30)) +
  scale_linetype_manual(values = c("Amoxicillin" =  "solid",
                                   "Ceftriaxone" = "solid",
                                   "Ciprofloxacin" = "solid",
                                   "Cotrimoxazole" = "solid",
                                   "Hospitalised" = "dashed",
                                   "TB therapy" = "dotted"),
                        breaks = c("Hospitalised",
                                   "TB therapy" ,
                                   "Ceftriaxone",
                                   "Cotrimoxazole", 
                                   "Ciprofloxacin", 
                                   "Amoxicillin")) + 
  scale_colour_manual(values = c("Amoxicillin" =  pal_lancet()(4)[1],
                                 "Ceftriaxone" = pal_lancet()(4)[2],
                                 "Ciprofloxacin" = pal_lancet()(4)[3],
                                 "Cotrimoxazole" = pal_lancet()(4)[4],
                                 "Hospitalised" = "black",
                                 "TB therapy" = "grey20"),
                      breaks = c("Hospitalised",
                                 "TB therapy" ,
                                 "Ceftriaxone",
                                 "Cotrimoxazole", 
                                 "Ciprofloxacin", 
                                 "Amoxicillin")) +
  theme(legend.title = element_blank()) + 
  xlab("Day post enrollment") + 
  ylab("Proportion") -> p1

outlong %>% ungroup() %>% select(-c(pid, assess_type, died, hosp,tb, cotri)) %>% 
  group_by(arm) %>% 
  summarise_all(sum) %>%
  pivot_longer(-arm) %>% group_by(name) %>% mutate(tots = sum(value)) %>% ungroup() %>%
  mutate(name = recode(name, !!!abs.lookup)) %>%
  ggplot(aes((fct_reorder(name, tots)), value)) + 
  geom_col() + 
  facet_wrap(~arm, scales = "free_x", labeller = as_labeller(arm_names)) +
  coord_flip() +
  theme_bw()  + 
  theme(panel.spacing = unit(1,"lines")) +
  ylab("Participant-days of exposure") +
  xlab(element_blank()) -> p2 

# cumulative fraction with no ab and no hosp
# start with hosp

abs.lookup[!names(abs.lookup) %in% c("tb","fluco", "ampho", "quin", "coart", "arte", "acicl", "flu", "hosp", "other")] -> anti_bacterials


apply(df.covs.cpt[c(names(anti_bacterials), "tb")], 1, sum) -> df.covs.cpt$any_antibacterial

df.covs.cpt$any_antibacterial[df.covs.cpt$any_antibacterial > 1] <- 1

time.to.ab.cot <- df.covs.cpt %>% group_by(pid) %>% do(ditch_everything_after_first_1(.,28))

#time.to.ab.cot <- filter(time.to.ab.cot, any_antibacterial > 0, arm == 3)

time.to.ab.cot %>%
  filter(any_antibacterial == 1) %>% 
  group_by(arm, assess_type) %>%
  tally() %>% group_by(arm) %>%
  mutate(cumsm = cumsum(n)) %>% 
  filter(arm == 2) %>% 
  ggplot(aes(assess_type, cumsm)) +
  geom_step(size = 0.75)  +  theme_bw() +
  ylab("Proportion") +
  xlab("Days post enrollment") +
  theme(legend.title = element_blank()) + ylim(c(0,30)) +
  scale_x_continuous(limits = c(0,120), breaks = c(0,30,60,90,120)) -> p3


ggarrange(ggarrange(NULL,p1,widths = c(0.09,1), ncol = 2, nrow = 1, labels = c("A", NA)),
          ggarrange(p2, NULL, nrow = 1, ncol = 2, widths = c(1,0.2), labels = c("B", NA)),
          ncol = 1, nrow = 2)


```

```{r exposure-tab, echo = F, warning = F, message = F}
outlong.cpt %>% ungroup() %>% 
  select(-c(assess_type, died)) %>% 
  pivot_longer(-c(pid, arm)) %>% 
  filter(value > 0) %>%
  group_by(pid, name,arm) %>% 
  summarise(pid.exp = sum(value)) %>% group_by(arm, name) %>%
  summarise(n.p = length(unique(pid)),
            exposure = sum(pid.exp),
            median.exp.len = median(pid.exp),
            lq.exp.len = quantile(pid.exp, 0.25),
            uq.exp.len = quantile(pid.exp, 0.75)) -> exposures
  
outlong.cpt %>% group_by(arm) %>%
  summarise(n.tot = length(unique(pid)),
            tot.exp = length(pid)) -> tots.exposures

exposures <- merge(tots.exposures, exposures)


exposures$exp.perc <- specify_decimal(exposures$exposure*100/exposures$tot.exp,0)
exposures$exp.perc[exposures$exp.perc == "0"] <- "<1"

exposures$exp.str <- as.character(exposures %>% glue_data('{exposure}')) #({exp.perc}%)'))
exposures$tr.len.str <- as.character(exposures %>% glue_data('{specify_decimal(median.exp.len, 0)} ({specify_decimal(lq.exp.len,0)}-{specify_decimal(uq.exp.len,0)})'))

exposures %>% select(name, arm,n.p, exp.str, tr.len.str) %>%
  pivot_wider(names_from = arm, 
              values_from = c(n.p, exp.str, tr.len.str),
              values_fill= list(n.p = 0, exp.str = "0", tr.len.str = "-")) -> exp.tab

#subset(exposures, arm == 1 & name != "hosp")
exp.tab$name <- factor(exp.tab$name, levels = c("hosp",
                                                subset(exposures, arm == 1 & name != "hosp")$name[order(subset(exposures, arm == 1 & name != "hosp")$exposure, decreasing = T)]))

exp.tab %>% mutate(name = recode(name, !!!abs.lookup)) -> exp.tab

exp.tab[order(exp.tab$name),] -> exp.tab
tots.exposures %>% pivot_wider(names_from = arm, values_from = c(n.tot, tot.exp)) -> tots.exposures
tots.exposures$name <- "Total At Risk"
names(tots.exposures)[1:6] <- names(exp.tab)[2:7]
tots.exposures[4:6] <- sapply(tots.exposures[4:6], as.character)
tots.exposures <- select(tots.exposures, names(exp.tab)[1:7] )
bind_rows(tots.exposures, exp.tab) -> exp.tab
exp.tab[is.na(exp.tab)] <- "-"

#exp.tab <- select(exp.tab, 
#                  n.p_1, exp.str_1, tr.len.str_1,
#                  n.p_2, exp.str_2, tr.len.str_2,
 #                 n.p_3, exp.str_3, tr.len.str_3)


kable(exp.tab, "latex", booktabs = TRUE,
      row.names = FALSE, escape = TRUE, col.names = c("Exposure", "Arm 1", "Arm 2", "Arm 3", "Arm 1", "Arm 2", "Arm 3", "Arm 1", "Arm 2", "Arm 3"),
      caption = "Antimicrobial and hospital exposure stratified by arm") %>% kable_styling(full_width = FALSE, latex_options = "scale_down") %>%
#  pack_rows("Total", 1,1) %>%
#  pack_rows("Hospitalisation", 2,2) %>%
 pack_rows("Exposures", 2,23) %>%
  add_header_above(c(" " = 1, "Number exposed" = 3, "Exposure (person-days)" = 3,
                     "Median (IQR) exposure length (days)" = 3)) %>% 
  footnote(general = "TB = tuberculosis, LA =lumefantrine artemether. Median exposure length includes only those exposed. Total at risk shows the total number of participants and participant-days of follow up included in the study." ,threeparttable = T, fixed_small_size = T)# %>% landscape()

```

```{r ESBL-a2-time-to-ab, echo = F, warning = F, message = F,fig.scap="Cumulative antibacterial exposure for baseline unexposed inpatients.", out.extra='', fig.cap = "Cumulative number of arm 2 particpants exposed to antibacterials (including CPT and TB therapy) as a function of time.",  fig.height = 2, fig.width=3, fig.align = 'center'}

p3

```

\clearpage

### ESBL-E colonisation

ESBL-E colonisation prevalence as a function of time across the three arms of the study is shown in Table \@ref(tab:ESBL-carriage-tab) and Figure \@ref(fig:ESBL-carriage-plots). Baseline colonisation prevalence was high in all groups, and higher in hospitalised participants than community members: 49% (95% CI 42-56%) in participants with sepsis, 41% (95% CI 32-52%) in baseline antimicrobial-unexposed inpatients and 28% (95% 20-38%) in community members. In crude analyses, both hospitalised groups show a rise in colonisation prevalence following admission, though this is much more marked in Arm 1 participants: by the day 7 visit 78% (95% CI 71-84%) of Arm 1 participants were colonised compared to 51% (38-64%) of Arm 2 participants.


```{r ESBL-carriage-tab, echo =F, message = F, warning = T}

merge(
lims %>% group_by(visit, arm) %>% 
  summarise(n = n(),
            e = sum(ESBL == "Positive"),
            prop = e/n,
            ESBLpropstr = as.character(glue('{e} ({specify_decimal(prop*100, 0)}%)')),
            nstr = as.character(n())) %>%
  select(visit, nstr, arm, ESBLpropstr) %>%
  pivot_wider(names_from = arm,values_from = c(ESBLpropstr,nstr), values_fill =list(ESBLpropstr = "-", nstr = "-")),

lims_orgs %>% group_by(visit, arm) %>% 
  mutate(n.smpl = length(unique(lab_id)),
         organism = recode(organism, `Escherichia coli` = "esco", `Klebsiella pneumoniae` = "klpn", .default = "other")) %>%
  filter(!is.na(organism)) %>% group_by(visit, arm, organism) %>% 
  summarise(n.sampl = unique(n.smpl),
            n = length(organism),
            prop = n/n.sampl,
            ORGpropstr = as.character(glue('{n} ({specify_decimal(prop*100, 0)}%)'))) %>%
  select(visit, arm, organism, ORGpropstr) %>%
  pivot_wider(names_from = c(arm,organism), values_from = ORGpropstr, values_fill = list( ORGpropstr = "-"))
) -> ESBLtab

ESBLtab <- select(ESBLtab, visit, 
                  nstr_1, ESBLpropstr_1,#`1_esco`,`1_klpn`, `1_other`,
                  nstr_2, ESBLpropstr_2,#`2_esco`,`2_klpn`, `2_other`,
                  nstr_3, ESBLpropstr_3)#`3_esco`,`3_klpn`, `3_other`)

ESBLtab$visit <- c("Day 0", "Day 7", "Day 28","Day 90", "Day 180")

kable(ESBLtab, "latex", booktabs = TRUE,
      row.names = FALSE, escape = TRUE, col.names = c("Visit",rep(c("n", "Any ESBL"),3)),
      caption = "ESBL carriage stratified by arm and visit") %>% kable_styling(full_width = FALSE) %>%
#  pack_rows("Total", 1,1) %>%
#  pack_rows("Hospitalisation", 2,2) %>%
# pack_rows("Exposures", 2,23) %>%
  add_header_above(c(" " = 1, "Arm 1 (Sepsis)" = 2, "Arm 2 (Inpatient)" = 2,
                     "Arm 3 (Community)" = 2))# %>% 
 # footnote(general = "TB = tuberculosis, LA =lumefantrine artemether. Median exposure length includes only those exposed. Total at risk shows the total number of participants and participant-days of follow up included in the study." ,threeparttable = T, fixed_small_size = T) %>% landscape()

```

```{r ESBL-carriage-plots, echo = F, warning = F, message = F,fig.scap="ESBL colonisation prevalence as a function of time", out.extra='', fig.cap = "ESBL carriage prevalence as a function of time visualised in a number of different ways. In each case participants from Arm 2 are censored on antimicrobial exposure and Arm 3 are censored on antimicrobial exposure or hospitalisation. Top (A) prevalence at each visit plotted at ostensible visit time; however, the visits are in fact distributed in time themselves so the middle plot (B) is an attempt to show this by fitting a nonparametric smoothed LOESS regression line with a local linear regression. However the confidence intervals in this method are too narrow because they assume indepence of the measurements, which are in fact clustered within patients. The bottom panel (C) is an estimate of the proportion of ESBL-colonised participants from the Aalen-Johansen estimate, which is a generalisation of the Kaplan-Meier curve. This takes into account the nonindependence of the measurements, but does not take into account the interval-censored nature of the data, and transitions to and from the ESBL colonised state are assumed to happen halfway between measurements. The best estimate of state occupancy to account for all these difficulties requires the fitting of a Markov model: see chapter xx.",  fig.height = 8, fig.width=6, out.height = '70%', fig.align = 'center'}


p.lci <- function(x,n) {
  return(binom.test(x,n)$conf.int[[1]])
}

p.uci <- function(x,n) {
  return(binom.test(x,n)$conf.int[[2]])
}



#ggplot(lims_dates, aes(assess_type, as.numeric(ESBL == "Positive"), group = arm)) + geom_point() + geom_smoo#th() + xlim(c(0, 200))

# first - plot at visit date

censor_list <- subset(time.to.ab.cot, any_antibacterial == 1 & arm != 1)

lims_dates_censored <- lims_dates

for (i in 1:nrow(censor_list)) {
  lims_dates_censored <- subset(lims_dates_censored, !(pid == censor_list$pid[i] & assess_type > censor_list$assess_type[i]))
}

lims_dates_censored %>% 
  group_by(arm, visit) %>%
  summarise(e = sum(ESBL == "Positive"),
            n = length(ESBL),
            prop = e/n,
            lci = p.lci(e,n),
            uci = p.uci(e,n) 
            ) %>% ungroup() %>% mutate(arm = as.factor(arm)) %>% 
  mutate(visit = recode(visit, `0` = 0, `1` = 7, `2` = 28, `3` = 90, `4` = 180),
         arm = recode(arm, `1` = "Arm 1", `2` = "Arm 2", `3` = "Arm 3")) %>%
  ggplot(aes(visit, prop, ymin = lci, ymax = uci, group = arm, fill = arm, color = arm)) +
  geom_line() + geom_ribbon(color = NA, alpha = 0.3) + theme_bw() +
  scale_color_manual(values = pal_lancet()(3)[c(2,1,3)]) + 
  scale_fill_manual(values = pal_lancet()(3)[c(2,1,3)]) +
  xlab("Day post enrollment") +
  ylab("Proportion ESBL") + coord_cartesian(ylim = c(0.2,0.9)) +
  theme(legend.title = element_blank()) -> p1.c

lims_dates_censored %>%
  mutate(arm = recode(arm, `1` = "Arm 1", `2` = "Arm 2", `3` = "Arm 3")) %>%
ggplot(aes(assess_type, as.numeric(ESBL == "Positive"),
           group = arm, 
           fill = as.factor(arm), 
           color = as.factor(arm ))) + 
  geom_smooth(size = 0.5) +
  coord_cartesian(xlim = c(0,180), ylim = c(0.2,0.9)) +
  theme_bw() +
  scale_color_manual(values = pal_lancet()(3)[c(2,1,3)]) + 
  scale_fill_manual(values = pal_lancet()(3)[c(2,1,3)]) +
  xlab("Day post enrollment") +
  ylab("Proportion ESBL") +
  theme(legend.title = element_blank()) -> p2.c



# both underestimate errors because assume independence of measurements
# fit a multistate model



#lims_dates %>% group_by(pid, visit, .drop = FALSE) %>% tally() %>% filter(n == 0)

#lims_dates_censored %>% unique %>% group_by(pid, visit) %>% tally() %>% filter(n > 1)

lims_dates_censored %>% filter(!pid %in% c("DAS1441S", 
                                  "DAS1497X",
                                  "DAS1519G")) %>%
  group_by(pid) %>% do(pointESBLmeas_tostartstop(.)) -> l.c

fit1.c <- survfit(Surv(time = tstart, time2 = tstop, event = ESBL, type = "mstate") ~ arm, data = l.c, id = pid, istate = init.state)

pull_state_occ_from_surv_obj(fit1) -> pstate.c

ggplot(pstate.c, aes(x = t, y = `p(ESBL.pos)`, colour = arm)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = `ll.ESBL.pos`, ymax = `ul.ESBL.pos`, fill = arm), colour = NA, alpha = 0.3) + 
  theme_bw() + 
  coord_cartesian(ylim = c(0.2,0.9), xlim = c(0,180))+
  labs(x = "Days post recruitment", y = "Proportion ESBL") -> p3.c

l.trans.halfway.c <- l.c %>% group_by(pid) %>% do(change_transition_times(., 0.5))
fit2.c <- survfit(Surv(time = tstart, time2 = tstop, event = ESBL, type = "mstate") ~ arm, data = l.trans.halfway.c, id = pid, istate = init.state)
pull_state_occ_from_surv_obj(fit2.c) -> pstate2.c


pstate2.c %>%
  mutate(arm = recode(arm, `1` = "Arm 1", `2` = "Arm 2", `3` = "Arm 3")) %>%
ggplot(aes(x = t, y = `p(ESBL.pos)`, colour = arm)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = `ll.ESBL.pos`, ymax = `ul.ESBL.pos`, fill = arm), colour = NA, alpha = 0.3) + 
  theme_bw() + 
  coord_cartesian(ylim = c(0.2,0.9), xlim = c(0,180)) +
  scale_color_manual(values = pal_lancet()(3)[c(2,1,3)]) + 
  scale_fill_manual(values = pal_lancet()(3)[c(2,1,3)]) +
  labs(x = "Days post recruitment", y = "Proportion ESBL") -> p4.c

ggarrange(p1.c,p2.c,p4.c, ncol = 1, nrow = 3, common.legend = TRUE, labels = c("A", "B", "C"))       

```

```{r ESL-car_cens}

```

### Determinents of baseline ESBL colonisation

### Determinents of d28 ESBL-E acquisition

## Discussion

### Limitations

## Conclusions and further work

